Modelling the gas kinematics of an atypical Lyman-alpha emitting compact dwarf galaxy

### 2017-10-04 paper/paper.tex
all models are wrong but some are useful

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
{\bf We argue that the results from the multiphase model 
provide a correct interpretation of the data.
In that case the clump velocity dispersion implies a dynamical mass of
$2\times 10^{9}$ M$_{\odot}$,  ten times its baryonic mass.}
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial galaxies, opening
{\bf the opportunity to use the models and methods presented in this
  paper to constrain the physics of star formation and feedback in the
  early generation of \lya emitting galaxies.} 
\end{abstract}

@@

\tol\ is a compact dwarf galaxy (CDG) that presents a
strong \lya emission with puzzling 
features: the line is highly symmetric, single peaked and broad 
\citep{Thuan97}.
{\bf The existence of this Symmetric Lyman Alpha Emitter (SLAE)} raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya radiation could emerge as a low surface
brightness glow, which may be connected to \lya halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep[see the review by][and references therein]{2014PASA...31...40D}.

@@

{\bf Motivated by observations of other CDGs
  that show gas kinematics ranging from pure rotation and low velocity
dispersion to high velocity dispersion without a clear rotation pattern
\citep{2015A&A...577A..21C,2017A&A...600A.125C} we perform here a new
study to explain \tol's \lya emission features under two physical
conditions for the interstellar medium: multiphase outflows and pure rotation.}

@@

Figure 1. Summarizes our main finding.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for the
multiphase model (thick line)  and the rotating homogeneous gas sphere
(thin line). 
The fit to the observations is not perfect. 
However, in spite of the
simplicity of our models, this is the first time that the main
features of {\bf this SLAE} can be reproduced: a broad, highly symmetric,
single-peaked \lya line.

@@

The low p-values are illustrated by the results shown in Figure \ref{multiphaseresults}.
Left column shows the difference between the integrated distributions
of the full sample (2500 input models) and the $1\%$ models with the lowest
$\chi^2/d.o.f.$, where we use the total number of degrees of freedom,
$d.o.f.=104$. 
The right column shows the actual $\chi^2/d.o.f.$ 
and its corresponding parameter value. 
Under these conditions we find $\sigma_{\rm cl}=$\sigmaclump,
$v_{\infty {\rm, cl}}=$\inftyclump\ and $P_{\rm cl}=$\probaclump, {\bf
  with the minimum $\chi^2/d.o.f.=11.7$.}

Qualitatively this result can be understood as follows.
Due to the large fraction of \lya photons being emitted within the
moving clumps ($P_{\rm cl}\sim 1$) the `intrinsic' profile closely follows 
the clump kinematics. 
In other words, the width of the intrinsic line
is set by $\sigma_{\mathrm{cl}}$ and its median offset mainly by
$v_{\infty,{\rm cl}}$ and $\beta_{\rm cl}$, {\bf the exponent defining
  the steepness of the radial velocity profile. } 
{\bf Furthermore, the relatively low mean number of clumps per line of
sight, $0.8<f_{\mathrm{cl}}<8$,} combined with the high velocity dispersion
of the clumps implies existence of low-density inter-clump 
regions where \lya photons can freely propagate.
This gives result to an emergent spectrum close to the intrinsic one, explaining the high
flux at the line's center.

@@

{\bf \tol\ presents a \lya emission line with puzzling features.}
Its flux at the line's center is high compared to other LAEs at low
redshift and the broad, highly symmetrical peak.
{\bf SLAEs are} virtually absent from other LAE surveys at low and high redshift
\citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. 
Simple shell models
fail to reproduce such a spectrum as reported by \cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models: multiphase ISM and solid
body rotation.

{\bf Which model could be closer to the actual kinematic conditions in \tol?  
Integral Field Unit (IFU) observations of other CDGs seem to
favor the multiphase model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial sampling was $0.67^{\prime\prime}$ and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine Blue Compact Galaxies (BCGs) to produce two-dimensional maps of the
continuum and prominent emission lines to finally compute velocity
field maps using the H$\alpha$ emission line. 
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120$ \kms) to highly irregular. 
The typical velocity dispersion values are in the range $10-50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms. }

{\bf These results disfavor the high rotational velocity of $V_{m}\sim350$
\kms\  that we find in the pure rotation model. 
On the other hand the results from the multiphase model yield a
velocity dispersion $\sigma_{\rm cl}=$\sigmaclump, consistent with observed
values in other CDGs, favoring the multiphase model.}
In this case we can  estimate a value for the dynamical
mass using the constraints for the velocity dispersion, $\sigma$,  in
a spherical system localized in a region of size $r$

@@

A larger dynamical mass estimate over its baryonic mass is a hint that
\tol\ is dark matter dominated.  
Following the methodology of \citet{2011ApJ...726..108T} we estimate
that a dark matter halo with a virial mass $\sim 6\times 10^{11}$
M$_{\odot}$ {\bf and a virial radius $220$ kpc } could explain this dynamical mass.
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.  
Figure \ref{fig:mass} shows the enclosed mass as a function of radius
for different dark matter halos together with the \tol's dynamical
mass estimate. 
{\bf This should be considered as an upper bound as lower values
  could be achieved if one considers instead a cored DM profile.}
{\bf Realistic mass estimates could only be achieved by detailed mass
  modelling once \tol's detailed kinematic information becomes
  available.}

{\bf New IFU observations are needed to confirm \tol's kinematics.
This could be done with the Multi Unit Spectroscopic Explorer (MUSE) 
\citep{2014Msngr.157...13B}, the Gemini Multi-Object Spectrographs (GMOS) 
\citep{2004PASP..116..425H} or the Fibre Large Array Multi Element
Spectrograph (FLAMES) \citep{2002Msngr.110....1P} as they have the spatial
resolution ($\sim0.5^{\prime\prime}$), spectral coverage and field
of view required to produce H$\alpha$ velocity maps to perform the
kind of study presented by \citet{2015A&A...577A..21C} on CDGs or by
\citet{Herenz16} on nearby \lya\ emitting galaxies.}

@@

All in all, the mere existence of {\bf a broad SLAE} is interesting.  
\tol's line shape is different to others and seems to reveal an
unexpected kinematic structure.
A confirmation of our results by new observations would support the multiphase model
as an element to be included in the study of \lya emitting
galaxies, {\bf consolidating the possibility to use the \lya line to
  constrain different models of star formation and feedback in the
  first generation of galaxies. }

>>>>

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
{\bf In this paper we reproduce for the first time these line features with two
different physically motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.}
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
We argue that the results from the multiphase model 
provide a correct interpretation of the data.
In that case the clump velocity dispersion implies a dynamical mass of
$2\times 10^{9}$ M$_{\odot}$,  ten times its baryonic mass.
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial galaxies, opening
the opportunity to use the models and methods presented in this
  paper to constrain the physics of star formation and feedback in the
  early generation of \lya emitting galaxies.
\end{abstract}

@@

\tol\ is a compact dwarf galaxy (CDG) that presents a
strong \lya emission with puzzling 
features: the line is highly symmetric, single peaked and broad 
\citep{Thuan97}.
The existence of this Symmetric Lyman Alpha Emitter (SLAE) raises the question whether some high
redshift LAEs have asymmetric lines because the blue half was
truncated by the intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya radiation could emerge as a low surface
brightness glow, which may be connected to \lya halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep[see the review by][and references therein]{2014PASA...31...40D}.

@@

Motivated by observations of other CDGs
  that show gas kinematics ranging from pure rotation and low velocity
dispersion to high velocity dispersion without a clear rotation pattern
\citep{2015A&A...577A..21C,2017A&A...600A.125C} we perform here a new
study to explain \tol's \lya emission features under two physical
conditions for the interstellar medium: multiphase outflows and pure rotation.

@@

Figure 1. Summarizes our main finding.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for the
multiphase model (thick line)  and the rotating homogeneous gas sphere
(thin line). 
The fit to the observations is not perfect. 
However, in spite of the
simplicity of our models, this is the first time that the main
features of this SLAE can be reproduced: a broad, highly symmetric,
single-peaked \lya line.

@@

The low p-values are illustrated by the results shown in Figure \ref{multiphaseresults}.
Left column shows the difference between the integrated distributions
of the full sample (2500 input models) and the $1\%$ models with the lowest
$\chi^2/d.o.f.$, where we use the total number of degrees of freedom,
$d.o.f.=104$. 
The right column shows the actual $\chi^2/d.o.f.$ 
and its corresponding parameter value. 
Under these conditions we find $\sigma_{\rm cl}=$\sigmaclump,
$v_{\infty {\rm, cl}}=$\inftyclump\ and $P_{\rm cl}=$\probaclump, 
  with the minimum $\chi^2/d.o.f.=11.7$. {\bf This relatively high value
    could be interpreted as low statistical significance.
However, we stress once more that this is the first time that the main
features of \tol\ can be reproduced, making this model an useful tool
to guide the interpretation of complex observational data.}

Qualitatively this result can be understood as follows.
Due to the large fraction of \lya photons being emitted within the
moving clumps ($P_{\rm cl}\sim 1$) the `intrinsic' profile closely follows 
the clump kinematics. 
In other words, the width of the intrinsic line
is set by $\sigma_{\mathrm{cl}}$ and its median offset mainly by
$v_{\infty,{\rm cl}}$ and $\beta_{\rm cl}$, the exponent defining
  the steepness of the radial velocity profile. 
Furthermore, the relatively low mean number of clumps per line of
sight, $0.8<f_{\mathrm{cl}}<8$, combined with the high velocity dispersion
of the clumps implies existence of low-density inter-clump 
regions where \lya photons can freely propagate.
This gives result to an emergent spectrum close to the intrinsic one, explaining the high
flux at the line's center.

@@

\tol\ presents a \lya emission line with puzzling features.
Its flux at the line's center is high compared to other LAEs at low
redshift and the broad, highly symmetrical peak.
SLAEs are virtually absent from other LAE surveys at low and high redshift
\citep{2012ApJ...751...29Y,LARS,Erb14,Trainor16}. 
Simple shell models
fail to reproduce such a spectrum as reported by \cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by two different kinematic models: multiphase ISM and solid
body rotation.

Which model could be closer to the actual kinematic conditions in \tol?  
Integral Field Unit (IFU) observations of other CDGs seem to
favor the multiphase model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial sampling was $0.67^{\prime\prime}$ and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine Blue Compact Galaxies (BCGs) to produce two-dimensional maps of the
continuum and prominent emission lines to finally compute velocity
field maps using the H$\alpha$ emission line. 
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120$ \kms) to highly irregular. 
The typical velocity dispersion values are in the range $10-50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms.

These results disfavor the high rotational velocity of $V_{m}\sim350$
\kms\  that we find in the pure rotation model. 
On the other hand the results from the multiphase model yield a
velocity dispersion $\sigma_{\rm cl}=$\sigmaclump, consistent with observed
values in other CDGs, favoring the multiphase model.
In this case we can  estimate a value for the dynamical
mass using the constraints for the velocity dispersion, $\sigma$,  in
a spherical system localized in a region of size $r$

@@

A larger dynamical mass estimate over its baryonic mass is a hint that
\tol\ is dark matter dominated.  
Following the methodology of \citet{2011ApJ...726..108T} we estimate
that a dark matter halo with a virial mass $\sim 6\times 10^{11}$
M$_{\odot}$ and a virial radius $220$ kpc could explain this dynamical mass.
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.  
Figure \ref{fig:mass} shows the enclosed mass as a function of radius
for different dark matter halos together with the \tol's dynamical
mass estimate. 
This should be considered as an upper bound as lower values
  could be achieved if one considers instead a cored DM profile.
Realistic mass estimates could only be achieved by detailed mass
  modelling once \tol's detailed kinematic information becomes
  available.

New IFU observations are needed to confirm \tol's kinematics.
This could be done with the Multi Unit Spectroscopic Explorer (MUSE) 
\citep{2014Msngr.157...13B}, the Gemini Multi-Object Spectrographs (GMOS) 
\citep{2004PASP..116..425H} or the Fibre Large Array Multi Element
Spectrograph (FLAMES) \citep{2002Msngr.110....1P} as they have the spatial
resolution ($\sim0.5^{\prime\prime}$), spectral coverage and field
of view required to produce H$\alpha$ velocity maps to perform the
kind of study presented by \citet{2015A&A...577A..21C} on CDGs or by
\citet{Herenz16} on nearby \lya\ emitting galaxies.

@@

All in all, the mere existence of a broad SLAE is interesting.  
\tol's line shape is different to others and seems to reveal an
unexpected kinematic structure.
A confirmation of our results by new observations would support the multiphase model
as an element to be included in the study of \lya emitting
galaxies, consolidating the possibility to use the \lya line to
  constrain different models of star formation and feedback in the
  first generation of galaxies.
